\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-07-12 papers/methods/paper.tex
Updating paper

% Abstract of the paper
\begin{abstract}
We present a hierarchical Bayesian model for use in supernova cosmology which takes into account a full range of analysis systematics and selection effects. This advances previous works with hierarchical Bayesian models by including additional sources of systematic uncertainty and improved treatment of Malmquist bias, whilst increasing numerical efficiency. Our model is optimised for efficient fitting and has been validated on a combined simulated dataset of more than $250\,000$ supernovae to rule out the presence of significant model biases in parameter recovery. Simplified statistical simulations for ideal supernovae cosmology show no bias, and sophisticated SNANA simulations of the combined low-redshift and DES 3-year spectroscopic survey sample quantify the bias for flat $w$CDM cosmology, with a matter prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$ included to allow rigorous determination of model bias in $w$. SNANA simulations can recover the input cosmology to within $0.27\sigma$, while taking into an increased number of systematics than prior analyses.
\end{abstract}
% Bias in SNANA tests across intrinsic dispersion models is found to range from $-0.15\sigma$ to $+0.27\sigma$ in $w$ for datasets of comparable strength to the Dark Energy Survey three-year analysis.

@@

\begin{table}
	\centering
	\caption{Investigating the combined 100 fits to {\gten} and {\celeven} simulations, fitting with both statistics only and also when including systematics. With systematics enabled, both the {\gten} and {\celeven} models show evidence of bias, scattering to either side of the simulated value of $w=-1$. However, their deviation from the truth value represents a shift of approximately $0.3\sigma$ when taking into account the uncertainty on fits to $w$. The bias is sub-dominant to both the size of the uncertainty for each fit, and the scatter induced by statistical variance in the simulations. }
	\label{tab:bulk_summary}
	\begin{tabular}{l|c}
		\hline
		Model & $w$ (scatter) \\
		\hline
		{\gten} Stat + Syst  &  $-1.03\pm 0.10\ (0.08)$   \\
		{\celeven} Stat + Syst  &  $-0.97\pm 0.09\ (0.06)$  \\
		{\gten}  Stat         &  $-1.01\pm 0.08\ (0.08)$   \\
		{\celeven} Stat         &  $-0.96\pm 0.07\ (0.05)$   \\
		\hline
	\end{tabular}
\end{table}

@@

 Combined posterior surfaces for 100 data realisations are shown in Figure \ref{fig:bulk_posterior}. Note that we have combined the posteriors for 100 realisations, and so we should expect the size of the uncertainty to be representative of one realisation, but the statistical spread of the final surface should be $\sqrt{100} = 10$ times less than a single realisation. The parameter bounds are listed in Table \ref{tab:bulk_summary}.

Cosmological bias is detected with $\approx 2.7\sigma$ significance for the {\celeven} model and $\approx 1.5\sigma$ significance for the {\gten} model when fitting both systematics and statistics. For the statistics only fits, the respective bias detection levels are $0.9\sigma$ and $5.5\sigma$. We investigate the {\celeven} bias and find its source to be a bias in the observed summary statistics, in addition to incorrect reported uncertainty on the summary statistics. This is confirmed by replacing the fully simulated observed $m_B$, $x_1$ and $c$ with random numbers drawn from a true Gaussian centered on the simulated SALT2 $m_B$, $x_1$ and $c$ values with covariance as reported by initial light curve fits. With this test, both the {\gten} and {\celeven} fits recover $w=-1.00$ exactly. From this, the main challenge of improving our methodology is to handle the fact that observational uncertainty is incorrect, non-Gaussian and biased. Our current model and techniques can quantify the effect of different scatter models, but being unable to constrain the `correct' (simulated) scatter model in our model fit means we cannot fully correct for the bias introduced an unknown scatter model.

>>>>

% Abstract of the paper
\begin{abstract}
We present a hierarchical Bayesian model for use in supernova cosmology which takes into account a full range of analysis systematics and selection effects. This advances previous works with hierarchical Bayesian models by including additional sources of systematic uncertainty and improved treatment of Malmquist bias, whilst increasing numerical efficiency. Our model is optimised for efficient fitting and has been validated on a combined simulated dataset of more than $250\,000$ supernovae to rule out the presence of significant model biases in parameter recovery. Simplified statistical simulations for ideal supernovae cosmology show no bias, and sophisticated SNANA simulations of the combined low-redshift and DES 3-year spectroscopic survey sample quantify the bias for flat $w$CDM cosmology, with a matter prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$ included to allow rigorous determination of model bias in $w$. SNANA simulations can recover the input cosmology to within $0.3\sigma$, while taking into an increased number of systematics than prior analyses.
\end{abstract}
% Bias in SNANA tests across intrinsic dispersion models is found to range from $-0.15\sigma$ to $+0.27\sigma$ in $w$ for datasets of comparable strength to the Dark Energy Survey three-year analysis.

@@

\begin{table}
	\centering
	\caption{Investigating the combined 100 fits to {\gten} and {\celeven} simulations, fitting with both statistics only and also when including systematics. The quoted value for $w$ represents the average mean of the fits, with the uncertainty being the average standard deviation of the fits. In brackets, the simulation scatter (the standard deviation of the mean of 100 fits) is shown, and the bias significance represents our confidence that the deviation in the mean $w$ away from $-1$ is not due to statistical fluctuation. With systematics enabled, both the {\gten} and {\celeven} models show evidence of bias, scattering to either side of the simulated value of $w=-1$. However, their deviation from the truth value represents a shift of approximately $0.3\sigma$ when taking into account the uncertainty on fits to $w$. The bias is sub-dominant to both the size of the uncertainty for each fit, and the scatter induced by statistical variance in the simulations. }
	\label{tab:bulk_summary}
	\begin{tabular}{l|cc}
		\hline
		Model & $w$ (scatter) & Bias Significance \\
		\hline
		{\gten} Stat + Syst  &  $-1.03\pm 0.10\ (0.08)$ & $3.9\sigma$  \\
		{\celeven} Stat + Syst  &  $-0.97\pm 0.09\ (0.06)$ & $4.4\sigma$  \\
		{\gten}  Stat         &  $-1.01\pm 0.08\ (0.08)$ & $2.0\sigma$   \\
		{\celeven} Stat         &  $-0.96\pm 0.07\ (0.05)$ & $7.9\sigma$   \\
		\hline
	\end{tabular}
\end{table}

@@

Points of maximum posterior for 100 data realisations are shown in Figure \ref{fig:bulk_posterior}. The parameter bounds are listed in Table \ref{tab:bulk_summary}.

Cosmological bias is detected with a confidence of more than $3\sigma$ for the {\gten} and {\celeven} models. We investigate the bias and find its source to be a bias in the observed summary statistics, in addition to incorrect reported uncertainty on the summary statistics. This is confirmed by replacing the fully simulated observed $m_B$, $x_1$ and $c$ with random numbers drawn from a true Gaussian centered on the simulated SALT2 $m_B$, $x_1$ and $c$ values with covariance as reported by initial light curve fits. With this test, both the {\gten} and {\celeven} fits recover $w=-1.00$ exactly. From this, the main challenge of improving our methodology is to handle the fact that observational uncertainty is incorrect, non-Gaussian and biased. Our current model and techniques can quantify the effect of different scatter models, but being unable to constrain the `correct' (simulated) scatter model in our model fit means we cannot fully correct for the bias introduced an unknown scatter model.
